---
title: "Hands-on Ex 02: Beyond ggplot2 Fundamentals"
author: "Li Ziyi"
---

## Library installation and launching

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
library(pacman)

pacman::p_load(tidyverse, 
               patchwork, 
               ggthemes, 
               hrbrthemes,
               ggrepel)
```

## Data Loading

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

## ggplot2 Annotation

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  geom_label(aes(label = ID), 
             hjust = .5, 
             vjust = -.5) +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")

```

To avoid text labels from being overlapping with each other, ggrepel is preferred here instead.

```{r}
ggplot(data=exam_data, 
       aes(x= MATHS, 
           y=ENGLISH)) +
  geom_point() +
  geom_smooth(method=lm, 
              linewidth=0.5) +  
  geom_label_repel(aes(label = ID), 
                   fontface = "bold") +
  coord_cartesian(xlim=c(0,100),
                  ylim=c(0,100)) +
  ggtitle("English scores versus Maths scores for Primary 3")
```

Content is yet to be released fully here, to be continued.
