---
title: "Hands-on Ex01: Creating Elegant Graphics with ggplot2"
author: "Li Ziyi"
---

# Getting Started

## Install and launch R packages

The code chunk below uses p_load() of pacman package to check if tidyverse packages are installed in the computer. If they are, then they will be launched into R.

```{r}
pacman::p_load(tidyverse)
```

## Importing the data

```{r}
exam_data <- read_csv("data/Exam_data.csv")
```

# Plotting a simple bar chart

```{r}

ggplot(data = exam_data,
       aes(x = RACE)) +
  geom_bar()
```

# Plotting a simple histogram

## Using R Graphics

```{r}

hist(exam_data$MATHS)
```

## Using ggplot2

```{r}

ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = 'green',
                 fill = 'navy') +
  ggtitle("Distribution of Maths scores")
```

## Using ggplot2 and group by genders

```{r}

ggplot(data = exam_data,
       aes(x = MATHS,
           fill = GENDER)) +
  geom_histogram(bins = 20,
                 boundary = 100,
                 color = 'grey30')
```

# Plotting a simple density estimate

```{r}

ggplot(data = exam_data,
       aes(x = MATHS,
           color = GENDER)) +
  geom_density()
```

# Plotting a simple boxplot with data points

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  geom_point(position = 'jitter',
             size = 0.5)
```

# Plotting a simple violin plot together with boxplot

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_violin(fill = 'light green') + 
  geom_boxplot(alpha = 0.25) +
  stat_summary(geom = 'point',
               fun.y = 'mean',
               color = 'maroon',
               size = 3)
```

# Plotting a simple scatter plot (maths vs english)

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = ENGLISH)) +
  geom_point(color = 'pink') +
  geom_smooth(size = 0.25)
```

# Plotting trellis using facet

```{r}

ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_wrap(~ CLASS)
```

# Plotting trellis using facet_grid

```{r}

ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20) +
  facet_grid(~ CLASS)
```

# Plotting trellis boxplot using facet_grid (slides 43)

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = CLASS)) +
  geom_boxplot(alpha = 0.5) +
  facet_grid(~ GENDER)
```

# Plotting trellis boxplot using facet_grid (slides 44)

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = CLASS)) +
  geom_boxplot(alpha = 0.5) +
  facet_grid(GENDER ~ .)
```

# Plotting trellis boxplot using facet_grid (slides 45)

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = GENDER)) +
  geom_boxplot() +
  facet_grid(GENDER ~ CLASS) 
```

# Plotting boxplot by fixing y&x range

```{r}

ggplot(data = exam_data,
       aes(y = MATHS,
           x = ENGLISH)) +
  geom_point() +
  geom_smooth(method = lm,
              size = 0.5) +
  coord_cartesian(xlim = c(0, 100),
                  ylim = c(0, 100)) + 
  theme_classic()
```

# Plotting barchart with requested theme (slide 53)

```{r}

ggplot(data = exam_data,
       aes(y = RACE)) +
  geom_bar() +
  theme_minimal() +
  theme(panel.background = element_rect(fill = 'light blue', color = 'white', size = 1, linetype = 'solid'),
        panel.grid.major = element_line(color = 'white', linetype = 'solid'),
        panel.grid.minor = element_line(color = 'white', linetype = 'solid'))
```

# Plotting barchart using requested format (slide 55)

```{r}

ggplot(data = exam_data,
       aes(x = reorder(RACE, RACE, function(x)-length(x)))) +
  geom_bar() +
  ylim(c(0,230))+
  geom_text(stat = 'count',
            aes(label = paste(after_stat(count), ', ', round(after_stat(count)/sum(count) * 100, 1), '%')),
            color = 'black',
            size = 3,
            vjust = -0.5) +
  xlab('Race') +
  ylab('No. of Pupils')
  # theme_minimal() +
  # theme(panel.background = element_rect(fill = 'grey80', color = 'white', size = 1, linetype = 'solid'),
  #       panel.grid.major = element_line(color = 'white', linetype = 'solid', linewidth = 0.5),
  #       panel.grid.minor = element_line(color = 'white', linetype = 'solid', linewidth = 0.5),
  #       axis.title.y = element_text(angle = 0))
```

# Plotting barchart using requested format (slide 59)

```{r}

ggplot(data = exam_data,
       aes(x = MATHS)) +
  geom_histogram(bins = 20,
                 fill = 'light blue',
                 color = 'black') + 
  geom_vline(aes(xintercept = mean(MATHS, na.rm = T),
                 color = 'red',
                 linetype = 'dashed'),
             show.legend = F) +
    geom_vline(aes(xintercept = median(MATHS, na.rm = T),
                 color = 'black',
                 linetype = 'dashed'),
             show.legend = F) +
  xlab('MATHS') +
  ylab('count')

```

# Plotting barchart for requested analysis (slide 60)

```{r}
#Don't know how to do this part... Remember to check prof's updated slides
ggplot(data = exam_data,
       aes(x = ENGLISH,
           fill = GENDER)) +
  geom_histogram() +
  facet_grid(~ GENDER) +
  geom_histogram(aes(x = ENGLISH),
                 alpha = 0.5) +  
  xlab('ENGLISH') +
  ylab('count')

```

------------------------------------------------------------------------
