---
title: "Hands-on Ex08: Modelling, visualising and analysing network data with R"
author: "Li Ziyi"
date: 3 March 2023
date-modified: "`r Sys.Date()`"
execute:
  echo: true
  eval: true
  warning: false
format: html
editor: visual
---

This hands-on exercise aims to

-   create graph object data frames, manipulate them using appropriate functions of dplyr, lubridate, and tidygraph,

-   build network graph visualisation using appropriate functions of ggraph,

-   compute network geometrics using tidygraph,

-   build advanced graph visualisation by incorporating the network geometrics, and

-   build interactive network visualisation using visNetwork package.

## Packages installation

```{r}
packages = c('igraph', 
             'tidygraph', 
             'ggraph', 
             'visNetwork', 
             'lubridate', 
             'clock',
             'tidyverse', 
             'graphlayouts')

for(p in packages){
  if(!require(p, character.only = T)){
    install.packages(p)
  }
  library(p, character.only = T)
}

```

## Dataset

The data to be used in this exercise is from an oil exploration and extraction company. There are two data sets. One contains the nodes data and the other contains the edges (also know as link) data.

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")

```

```{r}
glimpse(GAStech_edges)
```

Some data wrangling is needed. For example, notice that the *SentDate* is currently in character format. We will need to change it back to date data type.

## Data Wrangling

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

Both dmy() and wday() are functions of lubridate package. lubridate is an R package that makes it easier to work with dates and times. - dmy() transforms the SentDate to Date data type. - wday() returns the day of the week as a decimal number or an ordered factor if label is TRUE. The argument abbr is FALSE keep the data spelling in full, i.e. Monday. The function will create a new column in the data.frame i.e. Weekday and the output of wday() will save in this newly created field. - the values in the Weekday field are in ordinal scale.

```{r}
glimpse(GAStech_edges)
```

A close examination of GAStech_edges data frame reveals that it consists of individual e-mail flow records. This is not very useful for visualisation.

In view of this, we will aggregate the individual by date, senders, receivers, main subject and day of the week.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

```{r}
glimpse(GAStech_edges)
```

## Create network objects using tidygraph

In this section, we learn how to create a graph data model by using tidygraph package. It provides a tidy API for graph/network manipulation. While network data itself is not tidy, it can be envisioned as two tidy tables, one for node data and one for edge data. tidygraph provides a way to switch between the two tables and provides dplyr verbs for manipulating them. Furthermore it provides access to a lot of graph algorithms with return values that facilitate their use in a tidy workflow.

Two functions of tidygraph package can be used to create network objects, they are:

-   tbl_graph() creates a tbl_graph network object from nodes and edges data.

-   as_tbl_graph() converts network data and objects to a tbl_graph network.

-   activate() verb from tidygraph serves as a switch between tibbles for nodes and edges. All dplyr verbs applied to tbl_graph object are applied to the active tibble.

-   .N() function is used to gain access to the node data while manipulating the edge data.

-   .E() will give you the edge data.

-   .G() will give you the tbl_graph object itself.

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated, 
                           directed = TRUE)

GAStech_graph
```

From the output, notice that the node data is active. The notion of an active tibble within a tbl_graph object makes it possible to manipulate the data in one tibble at a time.

We can change which tibble data frame is active with the activate() function. Thus, if we wanted to rearrange the rows in the edges tibble to list those with the highest "weight" first, we could use activate() and then arrange().

```{r}
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## Plot network data with ggraph package

As in all network graph, there are three main aspects to a ggraph's network graph, they are: nodes, edges and layouts.

### Plot a basic network graph

The code chunk below uses ggraph(), geom-edge_link() and geom_node_point() to plot a network graph by using GAStech_graph.

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

Here, the basic plotting function is ggraph(), which takes the data to be used for the graph and the type of layout desired. Both of the arguments for ggraph() are built around igraph. Therefore, ggraph() can use either an igraph object or a tbl_graph object.

### Change the default network graph theme

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### Change the coloring of the plot

```{r}
g <- ggraph(GAStech_graph) + 
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

### Change the layout of ggraph

ggraph() support many layout for standard used, they are: star, circle, nicely (default), dh, gem, graphopt, grid, mds, spahere, randomly, fr, kk, drl and lgl.

```{r}
g <- ggraph(GAStech_graph, 
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### Change the corloring of the nodes

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department, 
                      size = 3))

g + theme_graph()
```

Take note that geom_node_point is equivalent in functionality to geo_point of ggplot2. It allows for simple plotting of nodes in different shapes, colours and sizes. In the codes chnuks above colour and size are used.

### Change the edges

We can map the thickness of edge to the Weight variable.

```{r}
g <- ggraph(GAStech_graph, 
            layout = "nicely") +
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 3)

g + theme_graph()
```

### Creating facet graphs

Another very useful feature of ggraph is faceting. In visualising network data, this technique can be used to reduce edge over-plotting in a very meaning way by spreading nodes and edges out based on their attributes. In this section, you will learn how to use faceting technique to visualise network data.

There are three functions in ggraph to implement faceting, they are:

-   facet_nodes() whereby edges are only draw in a panel if both terminal nodes are present here,

-   facet_edges() whereby nodes are always drawn in al panels even if the node data contains an attribute named the same as the one used for the edge facetting, and

-   facet_graph() faceting on two variables simultaneously.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)

g + facet_edges(~Weekday)
```
```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2) +
  theme(legend.position = 'bottom')
  
g + facet_edges(~Weekday)
```

```{r}
set_graph_style() 

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```

```{r}
set_graph_style()

g <- ggraph(GAStech_graph, 
            layout = "nicely") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department), 
                  size = 2)
  
g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",  
                border = TRUE) +
  theme(legend.position = 'bottom')
```
## Network metrics analysis
### Computing centrality indices
Centrality measures are a collection of statistical indices use to describe the relative important of the actors are to a network. There are four well-known centrality measures, namely: degree, betweenness, closeness and eigenvector. It is beyond the scope of this hands-on exercise to cover the principles and mathematics of these measure here. Students are encouraged to refer to Chapter 7: Actor Prominence of A User’s Guide to Network Analysis in R to gain better understanding of theses network measures.

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```
### Visualising network metrics
```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department, 
                      size = centrality_betweenness()))
g + theme_graph()
```

### Visualing community
tidygraph package inherits many of the community detection algorithms imbedded into igraph and makes them available to us, including Edge-betweenness (group_edge_betweenness), Leading eigenvector (group_leading_eigen), Fast-greedy (group_fast_greedy), Louvain (group_louvain), Walktrap (group_walktrap), Label propagation (group_label_prop), InfoMAP (group_infomap), Spinglass (group_spinglass), and Optimal (group_optimal). Some community algorithms are designed to take into account direction or weight, while others ignore it. Use this link to find out more about community detection functions provided by tidygraph,

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(group_edge_betweenness(weights = Weight, directed = TRUE))) %>%
  ggraph(layout = "fr") + 
  geom_edge_link(aes(width=Weight), 
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = community))  

g + theme_graph()
```

## Interactive network graph

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

```{r}
visNetwork(GAStech_nodes, 
           GAStech_edges_aggregated)
```
### To include a layout
```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") 

```

### To include visual attributes - nodes

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department) 


visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
### To include visual attributes - edges

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to", 
           smooth = list(enabled = TRUE, 
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
### Interactivity
visOptions() is used to incorporate interactivity features in the data visualisation.
-   The argument highlightNearest highlights nearest when clicking a node.
-   The argument nodesIdSelection adds an id node selection creating an HTML select element.
```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```











